---
title: "Filtering - Untargeted Metabolomics Data - Negative Mode"
author: "Giovana Domeneghini Mercali"
date: "2023-08-30"
format: gfm
---

# Loading libraries

```{r, message = FALSE}
library(tidyverse)
library(janitor) # if you want to clean_names()
```

Once you get deconvoluted data from MZmine or similar programs, you need to wrangle your data in such a way that you can conduct your analysis on it.

# Read in data

First we want to read in our raw data. The code here is to read in data directly from MZmine, so if you are using a different program for deconvolution, you might need to make some light adjustments.

```{r}
metadata_juice <- read_csv(file = "Feature_list_all_data_neg.csv",
                      col_names = TRUE, # has headers
                      na = "0")

# How many features do we have?
dim(metadata_juice)

# Look at beginning of the dataset

metadata_juice[1:8, 1:8]
# str(metadata_juice)
# names(metadata_juice)
#length(metadata_juice)
```

Note there is no metadata included in this file.  Just m/z, retention time, and a column for each sample, where values are peak heights. We are using peak height instead of peak area because it is less dependent on bad peak shape which you get sometimes with Metabolomics.

# RT filter 
You might have deconvoluted more data than you plan to use in your analysis. For example, you may want to exclude the first bit and last bit of your run, since you do not expect to have good reproducibility in those areas. 

Here, we are filtering to only include features that elute between 0.3-8.0 min of this 10 min run.

```{r}
metadata_juice_RTfilt <- metadata_juice %>%
  filter(between(`row retention time`, 0.3, 8.0))

# Did it work?
range(metadata_juice_RTfilt$`row retention time`)

# How many features do we have now?
dim(metadata_juice_RTfilt)
```
# Cleaning up data

## Create mz_rt

This creates a unique identifier for each feature using its mass-to-charge ratio (m/z) and retention time (RT).

```{r}
MZ_RT <- metadata_juice_RTfilt %>%
  mutate(mz = round(metadata_juice_RTfilt$`row m/z`, digits = 4), # decrease number of decimals for m/z & rt
         rt = round(metadata_juice_RTfilt$`row retention time`, digits = 4),
         .before = 1,
         .keep = "unused") %>%
  unite(mz_rt, c(mz, rt), remove = TRUE) %>% # Combine m/z & rt with _ in between
  rename(mz = `row m/z`,
         rt = `row retention time`,
         row_ID = `row ID`) %>% # rename to be more R friendly
  select(row_ID, everything()) # reorder and move row_ID to front

MZ_RT[1:8, 1:8]
```

## Clean up file names

We are using `gsub()` to replace strings (i.e. characters) in our sample names. Here is some useful info about [gsub](http://www.endmemo.com/r/gsub.php), and two different tutorials [here](https://www.youtube.com/watch?v=4ZokHoF99DY) and [here](https://www.youtube.com/watch?v=r4Sh7H6wzPA). You will likely need to change this code to suit your purposes.

```{r}
# remove stuff from the end of file names, ".mzML Peak height"
newcolumnnames <- gsub(".mzML.*","", colnames(MZ_RT))
colnames(MZ_RT) <- newcolumnnames
```

What are our sample names?
```{r}
colnames(MZ_RT)
```
## Remove unwanted samples

Removing conditioning QCs and solvent blanks.

```{r}
MZ_RT <- MZ_RT %>%
  select(-contains("Conditioning"), -contains("Solvent"))

colnames(MZ_RT)
```

# Start filtering

### CV function
Since base R does not have a function to calculate coefficient of variance, let's write one.
```{r}
cv <- function(x){
        (sd(x)/mean(x))
}
```


## Counting QCs

Subset QCs and filter features to keep only those that are present in 100% of QCs.  You could change this parameter based on your data. 

```{r}
# check dimensions of current df
dim(MZ_RT)

MZ_RT_QCs <- MZ_RT %>%
  select(mz_rt, contains("QC")) %>% # select QCs
  filter(rowSums(is.na(.)) <= 1) # remove rows that have 1 or more NAs
```

```{r}
# check dimensions of QCs filtered df
dim(MZ_RT_QCs)

# how many features got removed with this filtering?
nrow(MZ_RT) - nrow(MZ_RT_QCs)
```

## Filter on QC CV

Here we are removing features that have a CV of more than 30% in the QCs.  The rationale is that if a feature cannot be reproducibly measured in samples that are all the same, it should not be included in our analysis.

```{r}
# calculate CV row-wise (1 means row-wise)
QC_CV <- apply(MZ_RT_QCs[, 2:ncol(MZ_RT_QCs)], 1, cv)


# bind the CV vector back to the QC df
MZ_RT_QCs_CV <- cbind(MZ_RT_QCs, QC_CV)

# filter for keeping features with QC_CV <= 0.30 (or 30%)
MZ_RT_QCs_CVfilt <- MZ_RT_QCs_CV %>%
  filter(QC_CV <= 0.30)
```

How many features did I remove with this CV filtering?
```{r}
nrow(MZ_RT_QCs) - nrow(MZ_RT_QCs_CVfilt)
```

## Merge back the rest of the data

MZ_RT_QCs_CVfilt only contains the QCs, We want to keep only the rows that are present in this df, and then merge back all of the other samples present in MZ_RT.  We will do this by creating a vector that has the mz_rt features we want to keep, and then using `filter()` and `%in%` to keep only features that are a part of this list.
```{r}
dim(MZ_RT_QCs_CVfilt)
dim(MZ_RT)

# make a character vector of the mz_rt features we want to keep
# i.e., the ones that passed our previous filtering steps

features_to_keep <- as.character(MZ_RT_QCs_CVfilt$mz_rt)

MZ_RT_filt <- MZ_RT %>%
  filter(mz_rt %in% features_to_keep)

dim(MZ_RT_filt)
```

You should have the same number of features in MZ_RT_QCs_CVfilt as you do in your new filtered df MZ_RT_filt.

```{r}
all.equal(nrow(MZ_RT_QCs_CVfilt), nrow(MZ_RT_filt))
```


## Process blanks

We want to remove features that are present in our process blanks as they are not coming from compounds present in our samples.  In this dataset, the sample (there is only two, typically you would have at least 3 process blanks to average) representing this process blank (a sample that includes all the extraction materials, minus the sample, here the juice was replaced by mass with water) has "PB" in the sample name.

```{r}
# grab the name of the column/sample that is the process blank
grep("ProcessBlank", colnames(MZ_RT_filt), value = TRUE)
```

Calculate the average value across the QCs, then remove features that are not at least 10x higher in the QCs than in the process blank.  To do this we will use [`apply()`](https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/apply).

`apply(X, MARGIN, FUN,...)` where X is your df, MARGIN is 1 for row-wise, and 2 for col-wise, and FUN is your function

```{r}
# pull avg peak height across QCs
avg_height_QC <- apply(MZ_RT_QCs_CVfilt[, 2:ncol(MZ_RT_QCs_CVfilt)], 1, mean)

# bind back to rest of data
MZ_RT_filt_QC_avg <- cbind(MZ_RT_filt, avg_height_QC)

# check dimensions
dim(MZ_RT_filt_QC_avg)
```

Pull the name of your process blank, and make a new column that indicates how many fold higher your peak height is in your average QC vs your process blank.
```{r}
# pull name of process blank 
grep("ProcessBlank", colnames(MZ_RT_filt), value = TRUE)

# make a new column that has a value of how many fold higher peak height is in QCs as compared to PB
# then you can avg your PBs together and do the same thing
MZ_RT_filt_QC_avg$ProcessBlank1_neg_8[is.na(MZ_RT_filt_QC_avg$ProcessBlank1_neg_8)] <- 0
MZ_RT_filt_QC_avg$ProcessBlank2_neg_77[is.na(MZ_RT_filt_QC_avg$ProcessBlank2_neg_77)] <- 0

MZ_RT_filt_PB <- MZ_RT_filt_QC_avg %>% 
  mutate(avg_height_PB = (ProcessBlank2_neg_77 + ProcessBlank1_neg_8)/2) %>%
  mutate(fold_higher_in_QC = avg_height_QC/avg_height_PB) %>%
  select(row_ID, mz_rt, mz, rt, avg_height_QC, avg_height_PB, fold_higher_in_QC)

head(MZ_RT_filt_PB)
dim(MZ_RT_filt_PB)
```

We want to keep features that are at least 10x higher in QCs than process blanks, and we also want to keep Infs, because an Inf indicates that a feature absent in the process blanks (i.e., you get an Inf because you're trying to divide by zero).
```{r}
# keep features that are present at least 10x higher in QCs vs PB
# or, keep NAs because those are absent in blank
PB_features_to_keep <- MZ_RT_filt_PB %>%
  filter(fold_higher_in_QC > 10 | is.infinite(fold_higher_in_QC)) 

dim(PB_features_to_keep)
```

How many features did we remove?
```{r}
nrow(MZ_RT_filt_QC_avg) - nrow(PB_features_to_keep)
```

Wow removed a lot of garbage! This is great.

Bind back metdata.
```{r}
MZ_RT_filt_PBremoved <- MZ_RT_filt_QC_avg %>%
  filter(mz_rt %in% PB_features_to_keep$mz_rt)
```

## Duplicates

Do we have any duplicate features?
```{r}
get_dupes(MZ_RT_filt_PBremoved, mz_rt)
```

No duplicates found. In case there is duplicates, let's remove them. The code below is not currently running.

```{r, eval = FALSE}
MZ_RT_filt_PBremoved_nodupes <- MZ_RT_filt_PBremoved %>%
  distinct(mz_rt, .keep_all = TRUE)

# how many duplicates did we remove?
nrow(MZ_RT_filt_PBremoved) - nrow(MZ_RT_filt_PBremoved_nodupes)

# check for dupes again, should be none
get_dupes(MZ_RT_filt_PBremoved_nodupes, mz_rt)

dim(MZ_RT_filt_PBremoved_nodupes)
```

Remove samples that we don't need anymore.
```{r}
MZ_RT_filt_PBremoved_extraremoved <- MZ_RT_filt_PBremoved %>%
  select(-ProcessBlank2_neg_77, -ProcessBlank1_neg_8, -avg_height_QC, -FreshJuice_neq_17)

dim(MZ_RT_filt_PBremoved_extraremoved)
```
Separate samples MEF x Pressure.
```{r}
MZ_RT_filt_PBremoved_extraremoved_MEF <- MZ_RT_filt_PBremoved_extraremoved %>%
  select(-contains("HPP"), -contains("US"), -contains("UST"), -contains("70TT")) %>%
  select(-MEF_SS5_neg_36, -`40TT_10M1_neq_18`, -`40TT1_neg_53`, -`40TT2_neq_22` ,-`40TT3_neg_39`,-'40TT4_neg_49', -'40TT5_neg_61', -`40TT6_neg_69`)

dim(MZ_RT_filt_PBremoved_extraremoved_MEF)
glimpse(MZ_RT_filt_PBremoved_extraremoved_MEF)

MZ_RT_filt_PBremoved_extraremoved_pressure <- MZ_RT_filt_PBremoved_extraremoved %>%
  select(-contains("MEF"), -contains("MEF_SS"), -contains("40TT_10M")) %>% 
  select(-'70TT2_neq_25') 

dim(MZ_RT_filt_PBremoved_extraremoved_pressure)
glimpse(MZ_RT_filt_PBremoved_extraremoved_pressure)


```

# Save your file
Now you have a list of features present in your samples after filtering for CV in QCs, and removing all the extraneous columns we added to help us do this, along with removing any process blanks.
```{r, eval = FALSE}
write_csv(MZ_RT_filt_PBremoved_extraremoved,
          "Post_filtering_neg_2997.csv")

write_csv(MZ_RT_filt_PBremoved_extraremoved_MEF,
          "Post_filtering_neg_2997_MEF.csv")

write_csv(MZ_RT_filt_PBremoved_extraremoved_pressure,
          "Post_filtering_neg_2997_pressure.csv")

```

