---
title: "Metabolomics - Neg Data - MEF - Statistical analyses - Part II"
author: "Giovana Domeneghini Mercali"
prefer-html: true
format: gfm
---

# Load libraries
```{r, message = FALSE, warning = FALSE}
library(factoextra) # visualizing PCA results
library(glue) # for easy pasting
library(plotly) # quick interactive plots
library(proxyC) # more efficient large matrices
library(data.table) # easy transposing
library(janitor) # for cleaning names and checking duplicates
library(notame) # for collapsing ions coming from the same metabolite
library(doParallel) # for parallelizing notame specifically
library(patchwork) # for making multi-panel plots
library(readxl)
library(svglite)
library(pheatmap) # for heínsatmaps
library(ggridges) # for ridgeline plots
library(ggdist) # for nice dotplots
library(rstatix) # for pipeable stats testing
library(agricolae) # for posthoc tests 
library(ggpubr) # extension for adding stats to plots
library(rcompanion)
library(purrr)
library(multcompView) # Load the multcompView package for comparison letters

# this is at the end hoping that the default select will be that from dplyr
library(tidyverse) # for everything
```

# Read in data

Negative mode.

```{r}
# read in metabolomics data

feature_table_long_log2 <- read_csv("feature_table_long_log2__annotated_neg.csv",
                  trim_ws = TRUE) 

# read in meta data
metadata <- read_csv("metadata_neg_mef.csv",
                  trim_ws = TRUE)

# read in meta data
annotation <- read_csv("Annotation_neg.csv", 
                  locale=locale(encoding = "latin1"))

```
# Compariosn between compounds A

## Boxplots

Filter to select the compound you want to plot.

```{r}
compound_A <- feature_table_long_log2 %>%
  filter(id == "3,4-Dihydroxybenzoic acid" | id == "Gallic acid" | id == "Ferulic acid" | id == "Caffeic acid" | id =="2,4,6-Trihydroxybenzoic acid" | id == "Ellagic acid") %>%
   filter(sample_or_qc == "Sample")

```


Then plot.

```{r}
boxplot_compound_A <- compound_A %>%
  ggplot(aes(x = factor(treatment, levels = c("FJ", "MT", "MEF", "MEF+SS")), 
             y = rel_abund, fill = treatment)) +
  geom_boxplot(color = "black", alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.3) +
  facet_wrap(vars(id), nrow = 2, scales = "free_y") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5")) +
  labs(x = "Treatment",
  y = "log2(relative abundance)",
  title = "")

boxplot_compound_A


```



## Testing assumptions

### Normality


```{r}
# testing normality by group
compound_A %>%
  drop_na(rel_abund) %>% # remove NAs
  group_by(id) %>% # test by species
  shapiro_test(rel_abund) # test for normality
```

### Constant variance

```{r}
# testing constant variance

compound_A %>%
  drop_na(rel_abund) %>% # remove NAs
  levene_test(rel_abund ~ id) # test for constant variance
```

```{r}
compound_A %>%
  drop_na(rel_abund) %>%
  ggplot(aes(x = rel_abund, y = id, fill = id)) +
  ggridges::geom_density_ridges(alpha = 0.7) +
  # scale_y_discrete(labels = c("3,4-Dihydroxybenzoic acid", "Gallic acid", "Ferulic acid", "Caffeic acid", "2,4,6-Trihydroxybenzoic acid", "Ellagic acid")) +
  theme_minimal() +  
  theme(legend.position = "none") +
  labs(x = "Relative abundance",
       y = "Compound",
       title = "Distribution of relative abundace for some compounds in blended juice")
```


Find which features have no variance across treatments

```{r}
treatments_novariance <- compound_A %>%
  group_by(id) %>%
  summarize(stdev = sd(rel_abund),
            mean = mean(rel_abund)) %>%
  filter(stdev == 0)

```

No features with zero variance.


## Nonparametric tests

### kruskal Wallis test

```{r}
kruskal_compound_A <- compound_A %>%
  drop_na(rel_abund, treatment) %>%            # Remove NA values first
  group_by(id) %>%                             # Group by 'id'
  kruskal_test(rel_abund ~ treatment) %>%      # Apply the Kruskal-Wallis test
  add_significance() %>%     
  arrange(p)     
           
kruskal_compound_A
```


### Post-hoc test

Run Dunn test

```{r}
(kruskal_compound_A_posthoc <- compound_A %>%
      drop_na(rel_abund, treatment) %>%        # Remove NA values first
      group_by(id) %>%                         # Group by 'id'
      dunn_test(rel_abund ~ treatment,
                p.adjust.method = "BH"))
```

```{r}
# Create a column called comparison

kruskal_compound_A_posthocc_1 <- kruskal_compound_A_posthoc %>%
  mutate(comparison = glue("{group1} - {group2}")) 
  
# Group data by 'id' and create a nested data frame
group_cldList <- kruskal_compound_A_posthocc_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(p.adj ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()

# Display the result
group_cldList <- group_cldList %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))
  
```


Make a table that has the maximum relative abundance for each treatment.

```{r}
# Calculate maximum rel_abund for each id and treatment
max_relabund <- compound_A %>%
  drop_na(rel_abund, treatment) %>%  # Remove NA values
  group_by(id, treatment) %>%        # Group by 'id' and 'treatment'
  summarize(max_tot_relabund = max(rel_abund, na.rm = TRUE)) %>%  # Calculate max
  ungroup()  # Remove grouping

# Display the result
max_relabund
```


Bind the groups to the maximum relative abundance.

```{r}
dunn_compound_A_for_plotting <- max_relabund %>%
  left_join(group_cldList, by = c("id", "treatment"))

```


Plot

```{r}
boxplot_dunn <- boxplot_compound_A +
  geom_text(data = dunn_compound_A_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = dunn_compound_A_for_plotting$Letter)) +
  labs(caption = "For each compound, treatmnets with different letters are significant different unsing Krushal Wallis test")

boxplot_dunn 
```

## Parametric tests

### ANOVA test

```{r}
# Prepare your data 
clean_data <- compound_A %>%
  drop_na(rel_abund, treatment) %>%
  group_by(id) %>%
  nest()  # Nest data for each 'id'

# Apply ANOVA to each nested data frame
results <- clean_data %>%
  mutate(
    anova_model = map(data, ~ aov(rel_abund ~ treatment, data = .x)),
    anova_summary = map(anova_model, tidy)  # Use broom::tidy to summarize ANOVA results
  )

# Print results for each id
results %>%
  select(id, anova_summary) %>%
  unnest(anova_summary) %>%
  print()

# To perform Tukey HSD test and print results
results <- results %>%
  mutate(
    tukey_test = map(anova_model, TukeyHSD),
    tukey_summary = map(tukey_test, tidy)  # Use broom::tidy to summarize Tukey HSD results
  )

# Print Tukey HSD results
table_results <- results %>%
  select(id, tukey_summary) %>%
  unnest(tukey_summary) %>%
  arrange(adj.p.value)%>%
  print()
```

```{r}
# Create a column called comparison

table_results_1 <- table_results %>%
rename(comparison = `contrast`)

# Group data by 'id' and create a nested data frame
group_cldList_2 <- table_results_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(adj.p.value ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()


# Display the result
group_cldList_2 <- group_cldList_2 %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))

```

Join the groups to the maximum relative abundance.

```{r}
tukey_compound_A_for_plotting <- max_relabund %>%
  left_join(group_cldList_2, by = c("id", "treatment"))
```

```{r}
Boxplot_tukey_compound_A <- boxplot_compound_A +
  geom_text(data = tukey_compound_A_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = Letter)) +
  labs(title = "", tag = "D")

Boxplot_tukey_compound_A

```

```{r}
ggsave(plot = Boxplot_tukey_compound_A,
       filename = "boxplot_compound_A_neg.svg")
```


# Compariosn between compounds B

## Boxplots

Filter to select the compound you want to plot.

```{r}
compound_B <- feature_table_long_log2 %>%
  filter(id == "Epicatechin" |  id == "Procyanidin B2" | id == "Procyanidin C1" | id == "Phloretin" |id =="Phlorhizin" | id == "Quercitrin") %>%
   filter(sample_or_qc == "Sample")

```


Then plot.

```{r}
boxplot_compound_B <- compound_B %>%
  ggplot(aes(x = factor(treatment, levels = c("FJ", "MT", "MEF", "MEF+SS")), 
             y = rel_abund, fill = treatment)) +
  geom_boxplot(color = "black", alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.3) +
  facet_wrap(vars(id), nrow = 2, scales = "free_y") +
  theme_classic() +
  scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5")) +
  theme(legend.position = "none") +
  labs(x = "Treatment",
  y = "log2(relative abundance)",
  title = "")
  

```
## Testing assumptions

### Normality

```{r}
# testing normality by group
compound_B %>%
  drop_na(rel_abund) %>% # remove NAs
  group_by(id) %>% # test by species
  shapiro_test(rel_abund) # test for normality
```
### Constant variance

```{r}
# testing constant variance

compound_B %>%
  drop_na(rel_abund) %>% # remove NAs
  levene_test(rel_abund ~ id) # test for constant variance
```


```{r}
compound_B %>%
  drop_na(rel_abund) %>%
  ggplot(aes(x = rel_abund, y = id, fill = id)) +
  ggridges::geom_density_ridges(alpha = 0.7) +
  # scale_y_discrete(labels = c("Epicatechin", "Procyanidin B2", "Procyanidin C1", "Phloretin", "Phlorhizin", "Quercitrin")) +
  theme_minimal() +  
  theme(legend.position = "none") +
  labs(x = "Relative abundance",
       y = "Compound",
       title = "Distribution of relative abundace for some compounds in blended juice")
```

## Nonparametric tests

### kruskal Wallis test

```{r}
kruskal_compound_B <- compound_B %>%
  drop_na(rel_abund, treatment) %>%            # Remove NA values first
  group_by(id) %>%                             # Group by 'id'
  kruskal_test(rel_abund ~ treatment) %>%      # Apply the Kruskal-Wallis test
  add_significance() %>%     
  arrange(p)     
           
kruskal_compound_B
```

### Post-hoc test

Run Dunn test

```{r}
(kruskal_compound_B_posthoc <- compound_B %>%
      drop_na(rel_abund, treatment) %>%        # Remove NA values first
      group_by(id) %>%                         # Group by 'id'
      dunn_test(rel_abund ~ treatment,
                p.adjust.method = "BH"))
```


```{r}
# Create a column called comparison

kruskal_compound_B_posthocc_1 <- kruskal_compound_B_posthoc %>%
  mutate(comparison = glue("{group1} - {group2}")) 
  
# Group data by 'id' and create a nested data frame
group_cldList <- kruskal_compound_B_posthocc_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(p.adj ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()

# Display the result
group_cldList <- group_cldList %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))
  
```


Make a table that has the maximum relative abundance for each treatment.

```{r}
# Calculate maximum rel_abund for each id and treatment
max_relabund <- compound_B %>%
  drop_na(rel_abund, treatment) %>%  # Remove NA values
  group_by(id, treatment) %>%        # Group by 'id' and 'treatment'
  summarize(max_tot_relabund = max(rel_abund, na.rm = TRUE)) %>%  # Calculate max
  ungroup()  # Remove grouping

# Display the result
max_relabund
```


Bind the groups to the maximum relative abundance.

```{r}
dunn_compound_B_for_plotting <- max_relabund %>%
  left_join(group_cldList, by = c("id", "treatment"))

```

Plot

```{r}
boxplot_dunn <- boxplot_compound_B +
  geom_text(data = dunn_compound_B_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = dunn_compound_B_for_plotting$Letter)) +
  labs(caption = "For each compound, treatmnets with different letters are significant different unsing Krushal Wallis test")

boxplot_dunn 
```


## Parametric tests

### ANOVA test

```{r}
# Prepare your data 
clean_data <- compound_B %>%
  drop_na(rel_abund, treatment) %>%
  group_by(id) %>%
  nest()  # Nest data for each 'id'

# Apply ANOVA to each nested data frame
results <- clean_data %>%
  mutate(
    anova_model = map(data, ~ aov(rel_abund ~ treatment, data = .x)),
    anova_summary = map(anova_model, tidy)  # Use broom::tidy to summarize ANOVA results
  )

# Print results for each id
results %>%
  select(id, anova_summary) %>%
  unnest(anova_summary) %>%
  print()

# To perform Tukey HSD test and print results
results <- results %>%
  mutate(
    tukey_test = map(anova_model, TukeyHSD),
    tukey_summary = map(tukey_test, tidy)  # Use broom::tidy to summarize Tukey HSD results
  )

# Print Tukey HSD results
table_results <- results %>%
  select(id, tukey_summary) %>%
  unnest(tukey_summary) %>%
  arrange(adj.p.value)%>%
  print()
```

```{r}
# Create a column called comparison

table_results_1 <- table_results %>%
rename(comparison = `contrast`)

# Group data by 'id' and create a nested data frame
group_cldList_2 <- table_results_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(adj.p.value ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()


# Display the result
group_cldList_2 <- group_cldList_2 %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))

```

Join the groups to the maximum relative abundance.

```{r}
tukey_compound_B_for_plotting <- max_relabund %>%
  left_join(group_cldList_2, by = c("id", "treatment"))
```

```{r}
Boxplot_tukey_compound_B <- boxplot_compound_B +
  geom_text(data = tukey_compound_B_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = Letter)) +
  labs(title = "", tag = "A")

Boxplot_tukey_compound_B

```
```{r}
ggsave(plot = Boxplot_tukey_compound_B,
       filename = "boxplot_compound_B_neg.svg")
```


# Compariosn between compounds C

## Boxplots

Filter to select the compound you want to plot.

```{r}
compound_C <- feature_table_long_log2 %>%
  filter(id == "Coumaroyl quinic acid" |  id == "Feruloylquinic acid" | id == "Quercetin-3-glucoside" | id == "Sinapic acid" |id =="5-Caffeoylquinic acid" | id == "Naringin") %>%
   filter(sample_or_qc == "Sample")

```


Then plot.

```{r}
boxplot_compound_C <- compound_C %>%
  ggplot(aes(x = factor(treatment, levels = c("FJ", "MT", "MEF", "MEF+SS")), 
             y = rel_abund, fill = treatment)) +
  geom_boxplot(color = "black", alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.3) +
  facet_wrap(vars(id), nrow = 2, scales = "free_y") +
  theme_classic() +
  theme(legend.position = "none") +
 scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5")) +
  labs(x = "Treatment",
  y = "log2(relative abundance)",
  title = "")

```

## Testing assumptions

### Normality


```{r}
# testing normality by group
compound_C %>%
  drop_na(rel_abund) %>% # remove NAs
  group_by(id) %>% # test by species
  shapiro_test(rel_abund) # test for normality
```
### Constant variance

```{r}
# testing constant variance

compound_C %>%
  drop_na(rel_abund) %>% # remove NAs
  levene_test(rel_abund ~ id) # test for constant variance
```


```{r}
compound_C %>%
  drop_na(rel_abund) %>%
  ggplot(aes(x = rel_abund, y = id, fill = id)) +
  ggridges::geom_density_ridges(alpha = 0.5) +
  # scale_y_discrete(labels = c( "Coumaroyl quinic acid", "Feruloylquinic acid", "Quercetin-3-glucoside", "Sinapic acid", "5-Caffeoylquinic acid", "Naringin")) +
  theme_minimal() +  
  theme(legend.position = "none") +
  labs(x = "Relative abundance",
       y = "Compound",
       title = "Distribution of relative abundace for some compounds in blended juice")
```

## Nonparametric tests

### kruskal Wallis test

```{r}
kruskal_compound_C <- compound_C %>%
  drop_na(rel_abund, treatment) %>%            # Remove NA values first
  group_by(id) %>%                             # Group by 'id'
  kruskal_test(rel_abund ~ treatment) %>%      # Apply the Kruskal-Wallis test
  add_significance() %>%     
  arrange(p)     
           
kruskal_compound_C
```

### Post-hoc test

Run Dunn test

```{r}
(kruskal_compound_C_posthoc <- compound_C %>%
      drop_na(rel_abund, treatment) %>%        # Remove NA values first
      group_by(id) %>%                         # Group by 'id'
      dunn_test(rel_abund ~ treatment,
                p.adjust.method = "BH"))
```

```{r}
# Create a column called comparison

kruskal_compound_C_posthocc_1 <- kruskal_compound_C_posthoc %>%
  mutate(comparison = glue("{group1} - {group2}")) 
  
# Group data by 'id' and create a nested data frame
group_cldList <- kruskal_compound_C_posthocc_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(p.adj ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()

# Display the result
group_cldList <- group_cldList %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))
  
```


Make a table that has the maximum relative abundance for each treatment.

```{r}
# Calculate maximum rel_abund for each id and treatment
max_relabund <- compound_C %>%
  drop_na(rel_abund, treatment) %>%  # Remove NA values
  group_by(id, treatment) %>%        # Group by 'id' and 'treatment'
  summarize(max_tot_relabund = max(rel_abund, na.rm = TRUE)) %>%  # Calculate max
  ungroup()  # Remove grouping

# Display the result
max_relabund
```


Bind the groups to the maximum relative abundance.

```{r}
dunn_compound_C_for_plotting <- max_relabund %>%
  left_join(group_cldList, by = c("id", "treatment"))

```

Plot

```{r}
boxplot_dunn <- boxplot_compound_C +
  geom_text(data = dunn_compound_C_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = dunn_compound_C_for_plotting$Letter)) +
  labs(caption = "For each compound, treatmnets with different letters are significant different unsing Krushal Wallis test")

boxplot_dunn 
```


## Parametric tests

### ANOVA test

```{r}
# Prepare your data 
clean_data <- compound_C %>%
  drop_na(rel_abund, treatment) %>%
  group_by(id) %>%
  nest()  # Nest data for each 'id'

# Apply ANOVA to each nested data frame
results <- clean_data %>%
  mutate(
    anova_model = map(data, ~ aov(rel_abund ~ treatment, data = .x)),
    anova_summary = map(anova_model, tidy)  # Use broom::tidy to summarize ANOVA results
  )

# Print results for each id
results %>%
  select(id, anova_summary) %>%
  unnest(anova_summary) %>%
  print()

# To perform Tukey HSD test and print results
results <- results %>%
  mutate(
    tukey_test = map(anova_model, TukeyHSD),
    tukey_summary = map(tukey_test, tidy)  # Use broom::tidy to summarize Tukey HSD results
  )

# Print Tukey HSD results
table_results <- results %>%
  select(id, tukey_summary) %>%
  unnest(tukey_summary) %>%
  arrange(adj.p.value)%>%
  print()
```

```{r}
# Create a column called comparison

table_results_1 <- table_results %>%
rename(comparison = `contrast`)

# Group data by 'id' and create a nested data frame
group_cldList_2 <- table_results_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(adj.p.value ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()


# Display the result
group_cldList_2 <- group_cldList_2 %>%
  rename(treatment = Group) 
  
# mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))

```

Join the groups to the maximum relative abundance.

```{r}
tukey_compound_C_for_plotting <- max_relabund %>%
  left_join(group_cldList_2, by = c("id", "treatment"))
```

```{r}
Boxplot_tukey_compound_C <- boxplot_compound_C +
  geom_text(data = tukey_compound_C_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = Letter)) +
  labs(title = "", tag = "B")

Boxplot_tukey_compound_C

```


```{r}
ggsave(plot = Boxplot_tukey_compound_C,
       filename = "boxplot_compound_C_neg.svg")
```

# Compariosn between compounds E

 
 
## Boxplots

Filter to select the compound you want to plot.

```{r}
compound_E <- feature_table_long_log2 %>%
  filter(id == "3,4-Dihydroxybenzoic acid" | id == "2,4,6-Trihydroxybenzoic acid" | id == "Glucose" | id == "2,4,6-Tryhydroxybenzaldehyde") %>%
   filter(sample_or_qc == "Sample")

```


Then plot.

```{r}
boxplot_compound_E <- compound_E %>%
  ggplot(aes(x = factor(treatment, levels = c("FJ", "MT", "MEF", "MEF+SS")), 
             y = rel_abund, fill = treatment)) +
  geom_boxplot(color = "black", alpha = 0.8, outlier.shape = NA) +
  geom_jitter(height = 0, width = 0.3) +
  facet_wrap(vars(id), nrow = 2, scales = "free_y") +
  theme_classic() +
  theme(legend.position = "none") +
  scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5")) +
  labs(x = "Treatment",
  y = "log2(relative abundance)",
  title = "")

boxplot_compound_E

```

## Testing assumptions

### Normality


```{r}
# testing normality by group
compound_E %>%
  drop_na(rel_abund) %>% # remove NAs
  group_by(id) %>% # test by species
  shapiro_test(rel_abund) # test for normality
```
### Constant variance

```{r}
# testing constant variance

compound_E %>%
  drop_na(rel_abund) %>% # remove NAs
  levene_test(rel_abund ~ id) # test for constant variance
```


```{r}
compound_E %>%
  drop_na(rel_abund) %>%
  ggplot(aes(x = rel_abund, y = id, fill = id)) +
  ggridges::geom_density_ridges(alpha = 0.7) +
  # scale_y_discrete(labels = c("3,4-Dihydroxybenzoic acid", "2,4,6-Trihydroxybenzoic acid", "Glucose", "2,4,6-Trihydroxybenzaldehyde")) +
  theme_minimal() +  
  theme(legend.position = "none") +
  labs(x = "Relative abundance",
       y = "Compound",
       title = "Distribution of relative abundace for some compounds in blended juice")
```
 

## Nonparametric tests

### kruskal Wallis test

```{r}
kruskal_compound_E <- compound_C %>%
  drop_na(rel_abund, treatment) %>%            # Remove NA values first
  group_by(id) %>%                             # Group by 'id'
  kruskal_test(rel_abund ~ treatment) %>%      # Apply the Kruskal-Wallis test
  add_significance() %>%     
  arrange(p)     
           
kruskal_compound_E
```

### Post-hoc test

Run Dunn test

```{r}
(kruskal_compound_E_posthoc <- compound_E %>%
      drop_na(rel_abund, treatment) %>%        # Remove NA values first
      group_by(id) %>%                         # Group by 'id'
      dunn_test(rel_abund ~ treatment,
                p.adjust.method = "BH"))
```

```{r}
# Create a column called comparison

kruskal_compound_E_posthocc_1 <- kruskal_compound_E_posthoc %>%
  mutate(comparison = glue("{group1} - {group2}")) 
  
# Group data by 'id' and create a nested data frame
group_cldList <- kruskal_compound_C_posthocc_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(p.adj ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()

# Display the result
group_cldList <- group_cldList %>%
  rename(treatment = Group) 

 # mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))
  
```


Make a table that has the maximum relative abundance for each treatment.

```{r}
# Calculate maximum rel_abund for each id and treatment
max_relabund <- compound_E %>%
  drop_na(rel_abund, treatment) %>%  # Remove NA values
  group_by(id, treatment) %>%        # Group by 'id' and 'treatment'
  summarize(max_tot_relabund = max(rel_abund, na.rm = TRUE)) %>%  # Calculate max
  ungroup()  # Remove grouping

# Display the result
max_relabund
```

Bind the groups to the maximum relative abundance.

```{r}
dunn_compound_E_for_plotting <- max_relabund %>%
  left_join(group_cldList, by = c("id", "treatment"))

```

Plot

```{r}
boxplot_dunn <- boxplot_compound_E +
  geom_text(data = dunn_compound_E_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = dunn_compound_E_for_plotting$Letter)) +
  labs(caption = "For each compound, treatmnets with different letters are significant different unsing Krushal Wallis test")

boxplot_dunn 
```


## Parametric tests

### ANOVA test

```{r}
# Prepare your data 
clean_data <- compound_E %>%
  drop_na(rel_abund, treatment) %>%
  group_by(id) %>%
  nest()  # Nest data for each 'id'

# Apply ANOVA to each nested data frame
results <- clean_data %>%
  mutate(
    anova_model = map(data, ~ aov(rel_abund ~ treatment, data = .x)),
    anova_summary = map(anova_model, tidy)  # Use broom::tidy to summarize ANOVA results
  )

# Print results for each id
results %>%
  select(id, anova_summary) %>%
  unnest(anova_summary) %>%
  print()

# To perform Tukey HSD test and print results
results <- results %>%
  mutate(
    tukey_test = map(anova_model, TukeyHSD),
    tukey_summary = map(tukey_test, tidy)  # Use broom::tidy to summarize Tukey HSD results
  )

# Print Tukey HSD results
table_results <- results %>%
  select(id, tukey_summary) %>%
  unnest(tukey_summary) %>%
  arrange(adj.p.value)%>%
  print()
```

```{r}
# Create a column called comparison

table_results_1 <- table_results %>%
rename(comparison = `contrast`)

# Group data by 'id' and create a nested data frame
group_cldList_2 <- table_results_1 %>%
  group_by(id) %>%
  nest() %>%  # Nest data for each 'id'
  
# Apply cldList function on each nested data
mutate(cld_results = map(data, ~ cldList(adj.p.value ~ comparison, data = .x, threshold = 0.05))) %>%
 unnest(cld_results) %>%  # Unnest the results
  select(-data) %>%       # Remove the nested column and ungroup the result
  ungroup()


# Display the result
group_cldList_2 <- group_cldList_2 %>%
  rename(treatment = Group) 

#  mutate(treatment = str_replace(treatment, "FreshJuice", "Fresh Juice"))

```

Join the groups to the maximum relative abundance.

```{r}
tukey_compound_E_for_plotting <- max_relabund %>%
  left_join(group_cldList_2, by = c("id", "treatment"))
```

```{r}
Boxplot_tukey_compound_E <- boxplot_compound_E +
  geom_text(data = tukey_compound_E_for_plotting,
            aes(x = treatment,
                y = max_tot_relabund + 0.5,
                label = Letter)) +
  labs(caption = "Line represents the median abundance per treatment. 
       For each compound, treatmnets with different letters are significant different using Tukey's test",   
       title = "", tag = "E")

Boxplot_tukey_compound_E

```


```{r}
ggsave(plot = Boxplot_tukey_compound_E,
       filename = "boxplot_compound_E_neg.svg")
```

