---
title: "Metabolomic first pass data analysis - Negative ionization mode - MEF"
author: "Giovana Domeneghini Mercali"
format: gfm
---

### Load libraries
```{r, message = FALSE, warning = FALSE}
library(factoextra) # visualizing PCA results
library(glue) # for easy pasting
library(plotly) # quick interactive plots
library(proxyC) # more efficient large matrices
library(data.table) # easy transposing
library(janitor) # for cleaning names and checking duplicates
library(notame) # for collapsing ions coming from the same metabolite
library(doParallel) # for parallelizing notame specifically
library(patchwork) # for making multi-panel plots
library(rstatix) # for additional univariate functionality
library(readxl)
library(svglite)

# this is at the end hoping that the default select will be that from dplyr
library(tidyverse) # for everything
```

## Read in data

Negative mode.

```{r}
# read in metabolomics data
metab <- read_csv("Post_filtering_neg_2997_MEF.csv",
                  trim_ws = TRUE,
                  na = "0") # read in 0s to be NAs.

# read in meta data
metadata <- read_csv("metadata_neg_mef.csv",
                  trim_ws = TRUE)

```

Take a quick look at our data.

```{r}
# look at first 5 rows, first 5 columns 
metab[1:5,1:5]

# look at first 5 rows, all columns 
metadata[1:5,]

# check dimensions
dim(metab)
dim(metadata)
```


```{r}
# make metab sample columns all numeric
metab <- metab %>%
  mutate((across(.cols = 5:ncol(.), .fns = as.numeric)))

glimpse(metab)
```

## Wrangle sample names

Here, the samples are in columns and the features are in rows. Samples are coded so that the first number is the treatment code, and the last code is the run order. We are going to transpose the data so that samples are in rows and features are in columns, and we will also import the metadata about the samples.

```{r}
metab_t <- metab %>%
  select(-row_ID, -mz, -rt) %>%
  t() %>%
  as.data.frame() %>%
  rownames_to_column(var = "mz_rt")

# make the first row the column names
colnames(metab_t) <- metab_t[1,]

# then remove the first row and rename the column that should be called sample_name
metab_t <- metab_t[-1,] %>%
  rename(sample_name = mz_rt) %>%
  mutate((across(.cols = 2:ncol(.), .fns = as.numeric)))

metab_t[1:10, 1:10]

```

Add in the metadata and make a new column that will indicate whether a sample is a "sample" or a "QC". The metadata we have for the samples we are no longer using (process blanks etc) are removed.

```{r}

metab_plus <- left_join(metab_t, metadata, by = "sample_name") %>% # keeps only those samples in metab_t
  mutate(sample_or_qc = if_else(str_detect(sample_name, "QC"), true = "QC", false = "Sample")) %>%
  dplyr::select(sample_name, short_sample_name, treatment, run_order, soluble_solids, mass, sample_or_qc, everything()) %>% # and move metadata to the front
mutate((across(.cols = 5:6, .fns = as.numeric)))

```


Go from wide to long data.

```{r}

metab_plus_long <- metab_plus %>%
  pivot_longer(cols = -c(sample_name, short_sample_name, treatment, run_order, sample_or_qc, soluble_solids, mass),  # remove metadata
               names_to = "mz_rt",
               values_to = "rel_abund")

glimpse(metab_plus_long)

```

Also add separate columns for mz and rt, and making both numeric.

```{r}

metab_plus_long <- metab_plus_long %>%
  separate_wider_delim(cols = mz_rt,
                       delim = "_",
                       names = c("mz", "rt"),
                       cols_remove = FALSE) %>%
  mutate(across(.cols = 8:9, .fns = as.numeric))  %>% # convert mz and rt to numeric
  drop_na(mz, rt) # removing mz and rt values that are NA; maybe not actually necessary

# how did that go?
head(metab_plus_long)

```

## Correction - soluble solids and mass

```{r}

metab_corrected_long <- metab_plus_long |>
 mutate(rel_abund = if_else(!is.na(rel_abund), true = rel_abund*40/(soluble_solids*mass), false = rel_abund)) |>
  select(-soluble_solids, -mass, -sample_name) |>
  rename("sample_name" = short_sample_name)

head(metab_corrected_long)

```


## Data summaries

What mass range do I have?

```{r}
range(metab_corrected_long$mz)
```

What retention time range do I have?

```{r}
range(metab_corrected_long$rt)
```

How many samples are in each of my meta-data groups?

```{r}
# make wide data to make some calculations easier
metab_corrected_wide <- metab_corrected_long %>%
  dplyr::select(-mz, -rt) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund)

# by sample vs QC
metab_corrected_wide %>%
  count(sample_or_qc)
```

What does my data coverage across mz and rt look like?

```{r}
metab_corrected_long %>%
  group_by(mz_rt) %>% # so we only have one point per feature
  ggplot(aes(x = rt, y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Retention time (min)",
       y = "Mass to charge ratio (m/z)",
       title = "m/z by retention time plot (all features)",
       subtitle = "C18 reversed phase, negative ionization mode")
```

Distribution of masses

```{r}
metab_corrected_long %>%
  group_by(mz_rt) %>%
  ggplot(aes(x = mz)) +
  geom_histogram(bins = 100) +
  theme_minimal() +
  labs(x = "Mass to charge ratio (m/z)",
       y = "Number of features",
       title = "Distribution of features by mass")
```

Distribution of retention times

```{r}
metab_corrected_long %>%
  group_by(mz_rt) %>%
  ggplot(aes(x = rt)) +
  geom_density() +
  theme_minimal() +
  labs(x = "Retention time",
       y = "Number of features",
       title = "Distribution of features by retention time")
```


## Missing data

### Surveying missingness

How many missing values are there for each feature? In this dataset, missing values are coded as zero.

```{r fig.width=6, fig.height=3}

# all data including QCs
# how many missing values are there for each feature (row)
na_by_feature <- rowSums(is.na(metab)) %>%
  as.data.frame() %>%
  rename(missing_values = 1)


na_by_feature %>%
  ggplot(aes(x = missing_values)) +
  geom_histogram(bins = 67) + # since 67 samples
  theme_minimal() + 
  labs(title = "Number of missing values for each feature",
       x = "Number of missing values",
       y = "How many features have that many missing values")
```

How many features have no missing values?

```{r}
na_by_feature %>%
  count(missing_values == 0)
```

How many missing values are there for each sample?

```{r fig.width=6, fig.height=3}
# all data including QCs
# how many missing values are there for each feature (row)
na_by_sample <- colSums(is.na(metab)) %>%
  as.data.frame() %>%
  rename(missing_values = 1) %>%
  rownames_to_column(var = "feature") %>%
  filter(!feature %in% c("mz_rt", "rt", "mz", "row_ID"))

na_by_sample %>%
  ggplot(aes(x = missing_values)) +
  geom_histogram(bins = 100) + # there are 3473 features
  theme_minimal() + 
  labs(title = "Number of missing values for each sample",
       x = "Number of missing values",
       y = "How many samples have that many missing values")
```

Which features have a lot of missing values?

```{r}
contains_NAs_feature <- metab_corrected_long %>%
  group_by(mz_rt) %>%
  count(is.na(rel_abund)) %>%
  filter(`is.na(rel_abund)` == TRUE) %>% 
  arrange(desc(n))

head(contains_NAs_feature)
```

Which samples have a lot of missing values?

```{r}
contains_NAs_sample <- metab_corrected_long %>%
  group_by(sample_name) %>%
  count(is.na(rel_abund)) %>%
  filter(`is.na(rel_abund)` == TRUE) %>%
  arrange(desc(n))

head(contains_NAs_sample)
```


Are there any missing values in the QCs? (There shouldn't be.)

```{r}
metab_QC <- metab %>%
  dplyr::select(contains("QC"))

na_by_sample <- colSums(is.na(metab_QC)) %>%
  as.data.frame() %>%
  rename(missing_values = 1) %>%
  rownames_to_column(var = "feature") %>%
  filter(!feature == "mz_rt")

sum(na_by_sample$missing_values) # nope
```

### Imputing missing values

This is an optional step but some downstream analyses don't handle missingness well. Here we are imputing missing data with half the lowest value observed for that feature.

```{r}
# grab only the feature data and leave metadata
metab_corrected_wide_imputed <- metab_corrected_wide %>%
  dplyr::select(-c(1:4)) # the metadata columns

metab_corrected_wide_imputed[] <- lapply(metab_corrected_wide_imputed,
                                 function(x) ifelse(is.na(x), min(x, na.rm = TRUE)/2, x))

metab_corrected_wide_imputed_notame <- metab_corrected_wide_imputed

# bind back the metadata
metab_corrected_wide_imputed <- bind_cols(metab_corrected_wide[,1:4], metab_corrected_wide_imputed)
```


For notame later:

```{r}

data_notame <- bind_cols(metab_corrected_wide[,1], metab_corrected_wide_imputed_notame) %>%
 as.data.frame()
```


Did imputing work?

```{r}
# count missing values
metab_corrected_wide_imputed %>%
  dplyr::select(-c(1:4)) %>% # where the metadata is
  is.na() %>%
  sum()
```

Create long imputed dataset.

```{r}
metab_corrected_long_imputed <- metab_corrected_wide_imputed %>%
  pivot_longer(cols = 5:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund")

head(metab_corrected_long_imputed)
```

Let's also make separate mz and rt columns.

```{r}
metab_corrected_long_imputed <- metab_corrected_long_imputed %>%
  separate_wider_delim(cols = mz_rt,
                       delim = "_",
                       names = c("mz", "rt"),
                       cols_remove = FALSE)

metab_corrected_long_imputed$mz <- as.numeric(metab_corrected_long_imputed$mz)
metab_corrected_long_imputed$rt <- as.numeric(metab_corrected_long_imputed$rt)
```


## Feature clustering with `notame`

We want to cluster features that likely come from the same metabolite together, and we can do this using the package `notame`. You can learn more [here](http://127.0.0.1:24885/library/notame/doc/feature_clustering.html).

```{r, eval = FALSE}
browseVignettes("notame")
```

Let's make a m/z by retention time plot again before we start.

```{r}
(before_notame <- metab_corrected_long_imputed %>%
  group_by(mz_rt) %>% # so we only have one point per feature
  ggplot(aes(x = rt, y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Retention time (min)",
       y = "Mass to charge ratio (m/z)",
       title = "m/z by retention time plot before notame, 2997 features",
       subtitle = "C18 reverse phase, negative ionization mode"))
```

### Wrangling data

Transpose the wide data for notame and wrangle to the right format. Below is info from the documentation: 

* Data should be a data frame containing the abundances of features in each sample, one row per sample, each feature in a separate column
* Features should be a data frame containing information about the features, namely feature name (should be the same as the column name in data), mass and retention time


```{r}
# change to results to numeric
data_notame <- data_notame %>%
  mutate(across(-sample_name, as.numeric))

tibble(data_notame)
```

Create df with features.
```{r}
features <- metab_corrected_long_imputed %>%
  dplyr::select(mz_rt, mz, rt) %>%
  mutate(across(c(mz, rt), as.numeric)) %>%
  as.data.frame() %>%
  distinct()

glimpse(features)
class(features)
```

### Find connections

Set `cache = TRUE` for this chunk since its a bit slow especially if you have a lot of features (this step took ~10 min).

```{r, cache = TRUE}
connection <- find_connections(data = data_notame,
                               features = features,
                               corr_thresh = 0.9,
                               rt_window = 1/60,
                               name_col = "mz_rt",
                               mz_col = "mz",
                               rt_col = "rt")
```

```{r}
head(connection)
```

### Clustering

Now that we have found all of the features that are connected based on the parameters we have set, we now need to find clusters.

```{r}
clusters <- find_clusters(connections = connection, 
                          d_thresh = 0.8)
```

Assign a cluster ID to each feature to keep, and the feature that is picked is the one with the highest median peak intensity across the samples.

```{r}
# assign a cluster ID to all features
# clusters are named after feature with highest median peak height
features_clustered <- assign_cluster_id(data_notame, 
                                        clusters, 
                                        features, 
                                        name_col = "mz_rt")
```

Export out a list of your clusters this way you can use this later during metabolite ID.

```{r, eval = FALSE}
# export clustered feature list this way you have it
write_csv(features_clustered,
          "features_clustered_neg_MEF.csv")
```

Pull data out from the clusters and see how many features we removed/have now.

```{r}
# lets see how many features are removed when we only keep one feature per cluster
pulled <- pull_clusters(data_notame, features_clustered, name_col = "mz_rt")

cluster_data <- pulled$cdata
cluster_features <- pulled$cfeatures

# how many features did we originally have after filtering?
nrow(metab)

# how many features got removed during clustering?
nrow(metab) - nrow(cluster_features)

# what percentage of the original features were removed?
((nrow(metab) - nrow(cluster_features))/nrow(metab)) * 100
```

Reduce our dataset to include only our new clusters. `cluster_data` contains only the retained clusters, while `cluster_features` tells you also which features are a part of each cluster.

```{r}
# combined metadata_plus with cluster_features

metab_imputed_clustered_wide <- left_join(metab_corrected_wide_imputed[,1:4], cluster_data,
                                          by = "sample_name") 

dim(metab_imputed_clustered_wide) # we have 2474 features since 4 metadata columns

# make a long/tidy df
metab_imputed_clustered_long <- metab_imputed_clustered_wide %>%
  pivot_longer(cols = 5:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund") %>%
  separate_wider_delim(cols = mz_rt, # make separate columns for mz and rt too
                       delim = "_",
                       names = c("mz", "rt"),
                       cols_remove = FALSE) %>%
  mutate(across(.cols = c("mz", "rt"), .fns = as.numeric)) # make mz and rt numeric
```

Write out the final clustered dataset.

```{r}
write_csv(metab_imputed_clustered_long,
  "features_clustered_long__neg_MEF.csv")
```


Let's look at a m/z by retention time plot again after clustering.

```{r}
(after_notame <- metab_imputed_clustered_long %>%
  group_by(mz_rt) %>% # so we only have one point per feature
  ggplot(aes(x = rt, y = mz)) +
  geom_point() +
  theme_minimal() +
  labs(x = "Retention time (min)",
       y = "Mass to charge ratio (m/z)",
       title = "m/z by retention time plot after notame, 2289 features",
       subtitle = "C18 reverse phase, negative ionization mode"))
```

```{r, fig.width = 8, fig.height = 8}
before_notame / after_notame
```

## Assessing data quality

Let's make sure that our data is of good quality.

### Untransformed data

First we are going to convert the type of some of the columns to match what we want (e.g., run order converted to numeric, treatment to factor).

```{r}
tibble(metab_imputed_clustered_long)

# make run_order numeric
metab_imputed_clustered_long$run_order <- as.numeric(metab_imputed_clustered_long$run_order)

# make treatment and sample_or_qc a factor (i.e., categorical)
metab_imputed_clustered_long$treatment <- as.factor(metab_imputed_clustered_long$treatment)
metab_imputed_clustered_long$sample_or_qc <- as.factor(metab_imputed_clustered_long$sample_or_qc)

# did it work?
tibble(metab_imputed_clustered_long)
```

Let's make a boxplot to see how the metabolite abundance looks across different samples.

```{r}
metab_imputed_clustered_long %>%
  ggplot(aes(x = sample_name, y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90),
        legend.position = "none") +
  labs(title = "LC-MS (-) feature abundances by sample",
       subtitle = "Data is unscaled",
       x = "Sample",
       y = "Relative abundance")
```

Can't really see anything because data is skewed.

### Transformed data

#### Log2 transformed

We will log2 transform our data.

```{r}
metab_imputed_clustered_long_log2 <- metab_imputed_clustered_long %>%
  mutate(rel_abund = log2(rel_abund))
```

And then plot.

```{r}
boxplot_log2_transformed <- metab_imputed_clustered_long_log2 %>%
  ggplot(aes(x = sample_name, y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5",
                                "QC" = "#F6A900")) +
  labs(y = "Relative abundance", x = "Sample", 
  title = "LC-MS (-) feature abundances by Sample",
       subtitle = "Data is log2 transformed")

boxplot_log2_transformed 
```


We can also look at this data by run order to see if we have any overall run order effects visible.

```{r}
boxplot_log2_transformed_run_order <- metab_imputed_clustered_long_log2 %>%
  mutate(sample_name = fct_reorder(sample_name, run_order)) %>%
  ggplot(aes(x = sample_name , y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_fill_manual(values = c("FJ" = "#F28EAD", 
                                "MT" = "#C77CDB",
                                "MEF" = "#A8D8B9", 
                                "MEF+SS" = "#7FB3D5",
                                "QC" = "#F6A900")) +
  labs(y = "Relative abundance", x = "Sample")

boxplot_log2_transformed_run_order
```


#### Log10 transformed

We will log10 transform our data.

```{r}
metab_imputed_clustered_long_log10 <- metab_imputed_clustered_long %>%
  mutate(rel_abund = log10(rel_abund))
```

We can look at this data where we group by species. 

```{r}
metab_imputed_clustered_long_log10 %>%
  ggplot(aes(x = sample_name , y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "LC-MS (-) feature abundances by sample",
       subtitle = "Data is log10 transformed",
       y = "Relative abundance", x = "Sample")
```

We can also look at this data by run order to see if we have any overall run order effects visible.

```{r}
metab_imputed_clustered_long_log10 %>%
  mutate(sample_name = fct_reorder(sample_name, run_order)) %>%
  ggplot(aes(x = sample_name , y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "LC-MS (+) feature abundances by sample",
       subtitle = "Data is log10 transformed",
       y = "Relative abundance", x = "Sample")
```

#### Autoscaled

Scales to unit variance, where the mean for each feature is 0 with a standard deviation of 1. This works well when all metabolites are considered to be equivalently important though measurement errors can be inflated. We've never actually scaled data this way for a project but whatever here it is.

```{r}
metab_imputed_clustered_wide <- metab_imputed_clustered_long %>%
  select(-mz, -rt) %>%
  pivot_wider(names_from = "mz_rt",
              values_from = "rel_abund")

# autoscaling on the now zero-centered matrix
autoscaled <- 
    bind_cols(metab_imputed_clustered_wide[1:4], # metadata
            lapply(metab_imputed_clustered_wide[5:ncol(metab_imputed_clustered_wide)], # metab data
                   scale)) # scale to mean 0 sd 1

autoscaled[1:10,1:10]

```

```{r}
autoscaled_long <- autoscaled %>%
  pivot_longer(cols = 5:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund") %>%
  mutate(treatment = as.factor(treatment))

autoscaled_long %>%
  mutate(sample_name = fct_reorder(sample_name, run_order)) %>%
  ggplot(aes(x = sample_name , y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "LC-MS (-) Feature Abundances by Sample",
       subtitle = "Data is autoscaled",
       y = "Relative abundance")
```

This is weird we won't use this.

#### Pareto scaled

Pareto scaling scales but keeps the fidelity of the original differences in absolute measurement value more than autoscaling. Often data is Pareto scaled after log transformation

```{r pareto scale}
metab_imputed_clustered_wide_log2 <- metab_imputed_clustered_long_log2 %>%
  select(-mz, -rt) %>%
  pivot_wider(names_from = "mz_rt",
              values_from = "rel_abund")

metab_imputed_clustered_wide_log2_metabs <- 
  metab_imputed_clustered_wide_log2[,5:ncol(metab_imputed_clustered_wide_log2)]

pareto_scaled <- 
  IMIFA::pareto_scale(metab_imputed_clustered_wide_log2_metabs, center = FALSE)

pareto_scaled <- bind_cols(metab_imputed_clustered_wide_log2[,1:4], pareto_scaled)
```



```{r}
pareto_scaled_long <- pareto_scaled %>%
  pivot_longer(cols = 5:ncol(.),
               names_to = "mz_rt",
               values_to = "rel_abund")

pareto_scaled_long %>%
  # mutate(short_sample_name = fct_reorder(short_sample_name, treatment)) %>%
  ggplot(aes(x = sample_name, y = rel_abund, fill = treatment)) +
  geom_boxplot(alpha = 0.6) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(title = "LC-MS (-) feature abundances by sample",
       subtitle = "Data is Pareto scaled",
       y = "Relative abundance", x = "Sample")
```

I think pareto scaling is making everything look super the same. I am going to use log2 transformed data for the rest of this analysis.

Let's write out that data as our final feature table used for the rest of the analysis.

```{r}
metab_imputed_clustered_wide_log2 <- metab_imputed_clustered_long_log2 %>%
  select(-mz, -rt) %>%
  pivot_wider(names_from = mz_rt,
              values_from = rel_abund)

write_csv(metab_imputed_clustered_wide_log2,
          file = "features_clustered_wide_log2_neg_MEF.csv")
```


## PCAs all data

### With QCs

```{r}
pca_qc <- prcomp(metab_imputed_clustered_wide_log2[,-c(1:4)], # remove metadata
                 scale = FALSE, # we did our own scaling
                 center = TRUE) # true is the default

summary(pca_qc)
```

Look at how much variance is explained by each PC.
```{r}
importance_qc <- summary(pca_qc)$importance %>%
  as.data.frame()

head(importance_qc)
```

Generate a scree plot.
```{r}
fviz_eig(pca_qc)
```

Generate a scores plot (points are samples) quickly with `fviz_pca_ind`.
```{r}
fviz_pca_ind(pca_qc)
```

Make a scores plot but prettier.
```{r}
# create a df of pca_qc$x
scores_raw_qc <- as.data.frame(pca_qc$x)

# bind meta-data
scores_qc <- bind_cols(metab_imputed_clustered_wide_log2[,1:4], # first 4 columns
                       scores_raw_qc)
```

Plot.
```{r}
# create objects indicating percent variance explained by PC1 and PC2
PC1_percent_qc <- round((importance_qc[2,1])*100, # index 2nd row, 1st column, times 100
                         1) # round to 1 decimal
PC2_percent_qc <- round((importance_qc[2,2])*100, 1) 

# plot
# aes(text) is for setting tooltip with plotly later to indicate hover text
(scores_qc_plot <- scores_qc %>%
  ggplot(aes(x = PC1, y = PC2, fill = treatment, text = glue("Sample: {sample_name},
                                                           Treatment: {treatment}"))) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(shape = 21, color = "black") +
  theme_minimal() +
  labs(x = glue("PC1: {PC1_percent_qc}%"), 
       y = glue("PC2: {PC2_percent_qc}%"), 
       title = "PCA scores plot colored by treatment, negative mode"))
```

Then make your scores plot ineractive so you can see who is who.

```{r eval = FALSE}
ggplotly(scores_qc_plot, tooltip = "text")
```

Make a loadings plot (points are features) even though it might not be that useful.

```{r}
fviz_pca_var(pca_qc)
```

See what I mean? Not that useful. There are some functions in PCAtools that label only the points that most contribute to each PC. Could also do this manually if its of interest.

### Without QCs

```{r}
metab_imputed_clustered_wide_log2_noqc <- metab_imputed_clustered_wide_log2 %>%
  filter(sample_or_qc == "Sample")


pca_noqc <- prcomp(metab_imputed_clustered_wide_log2_noqc[,-c(1:4)], # remove metadata
                 scale = FALSE, # we did our own scaling
                 center = TRUE) # true is the default

summary(pca_noqc)
```

Look at how much variance is explained by each PC.

```{r}
importance_noqc <- summary(pca_noqc)$importance %>%
  as.data.frame()

head(importance_noqc)
```

Generate a scree plot.

```{r}
fviz_eig(pca_noqc)
```

Generate a scores plot (points are samples) quickly with `fviz_pca_ind`.

```{r}
fviz_pca_ind(pca_noqc)
```

Make a scores plot but prettier.

```{r}
# create a df of pca_qc$x
scores_raw_noqc <- as.data.frame(pca_noqc$x)

# bind meta-data
scores_noqc <- bind_cols(metab_imputed_clustered_wide_log2_noqc[,1:4], # metadata
                         scores_raw_noqc)
```

Plot.

```{r}
# create objects indicating percent variance explained by PC1 and PC2
PC1_percent_noqc <- round((importance_noqc[2,1])*100, # index 2nd row, 1st column, times 100
                         1) # round to 1 decimal
PC2_percent_noqc <- round((importance_noqc[2,2])*100, 1) 

# plot
# aes(text) is for setting tooltip with plotly later to indicate hover text
(scores_noqc_plot <- scores_noqc %>%
  ggplot(aes(x = PC1, y = PC2, fill = treatment, text = glue("Sample: {sample_name},
                                                             Treatment: {treatment}"))) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point(shape = 21, color = "black") +
  theme_minimal() +
  labs(x = glue("PC1: {PC1_percent_noqc}%"), 
       y = glue("PC2: {PC2_percent_noqc}%"), 
       title = "PCA scores plot colored by treatment, negative mode"))
```

Then make your scores plot ineractive so you can see who is who.

```{r eval = FALSE}
ggplotly(scores_noqc_plot, tooltip = "text")
```

Make a loadings plot (points are features) even though it might not be that useful.

```{r}
fviz_pca_var(pca_noqc)
```

See what I mean? Not that useful. There are some functions in PCAtools that label only the points that most contribute to each PC. Could also do this manually if its of interest.

